---
title: "Getting started with flowcam"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with flowcam}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

**flowcam** provides a tidy interface to the USGS [National Imagery Management System (NIMS)](https://api.waterdata.usgs.gov/nims/v0) API, which serves stream-gage camera images collected across the United States.

## Authentication

NIMS requests can be made without an API key, but unauthenticated calls are rate-limited. Register for a free key at <https://api.waterdata.usgs.gov/signup/> and store it with `set_nims_key()`:

```{r auth}
library(flowcam)

set_nims_key("your_api_key_here")
```

`set_nims_key()` writes `API_USGS_PAT` to `~/.Renviron` and applies it to the current session immediately. All subsequent API calls in the session—and in future R sessions—use that key automatically. The same environment variable is read by the `dataRetrieval` package, so one key covers both.

## Discovering cameras

`find_cameras()` returns a tibble of camera metadata. With no arguments it fetches all available cameras; pass `site_id` to narrow results to a specific NWIS site.

The Pecos Web Camera near Roswell, NM (NWIS site 08385630) is a well-known site on the Pecos River:

```{r find-cameras}
cam <- find_cameras(site_id = "08385630")
cam
```

Key columns returned:

| Column | Description |
|---|---|
| `camId` | Camera identifier used in all other functions |
| `camName` | Human-readable name |
| `lat`, `lng` | Decimal-degree location |
| `newestImageDT` | Timestamp of the most recent image (UTC) |
| `TL_enabled` | Whether daily timelapse videos are generated |
| `smallDir`, `overlayDir`, `thumbDir` | S3 base directories for each image size |

If you already know a camera ID, look it up directly:

```{r find-by-cam-id}
find_cameras(cam_id = cam$camId[[1]])
```

To limit network traffic when you only need a few fields, use `return_fields`:

```{r return-fields}
find_cameras(
  site_id       = "08385630",
  return_fields = c("camName", "newestImageDT", "smallDir")
)
```

## Listing available images

`list_images()` returns the filenames stored for a camera. Filenames alone are enough to build download URLs via `build_image_url()`.

```{r list-images}
cam_id <- cam$camId[[1]]

# 10 most recent images
imgs <- list_images(cam_id, limit = 10)
imgs
```

By default the result is a single-column tibble of filenames. Set `raw_item = TRUE` to also retrieve timestamps and file sizes:

```{r list-raw}
list_images(cam_id, limit = 10, raw_item = TRUE)
```

### Filtering by date and time

Pass POSIXct values (in UTC) to `after` and `before` to fetch images from a specific window:

```{r date-filter}
list_images(
  cam_id,
  after  = as.POSIXct("2025-06-01 00:00:00", tz = "UTC"),
  before = as.POSIXct("2025-06-02 00:00:00", tz = "UTC")
)
```

## Building image URLs

`build_image_url()` combines the camera's base S3 directory with the filenames returned by `list_images()`:

```{r build-urls}
urls <- build_image_url(cam, imgs$filename, size = "small")
head(urls)
```

The `size` argument controls image dimensions:

- `"small"` (default) — ~720 px wide, suitable for monitoring dashboards
- `"overlay"` — full-size overlay image with gage-reading annotation
- `"thumb"` — thumbnail ~200 px tall

## Downloading images

`download_images()` wraps the steps above into a single call: it finds the camera, lists images, builds URLs, and saves the files to a local directory.

```{r download}
dest <- tempdir()

paths <- download_images(
  cam_id   = cam_id,
  dest_dir = dest,
  size     = "small",
  limit    = 5
)

paths
```

Already-downloaded files are skipped by default (`overwrite = FALSE`), making it easy to resume a partially completed download. Pass `overwrite = TRUE` to force re-download.

## Timelapse videos

Many USGS cameras produce a daily timelapse MP4 stitched from images collected throughout the day. `get_timelapse_url()` returns the URL to that file:

```{r timelapse}
url <- get_timelapse_url(cam_id)
url
```

You can open the URL directly in a browser or pass it to a video tool. A warning is emitted when `TL_enabled` is `FALSE` for the camera.

## Next steps

- See `vignette("pecos-river")` for a full monitoring workflow combining **flowcam** with streamflow data from **dataRetrieval**.
- Use `find_gage_cameras()` to retrieve camera metadata enriched with NWIS site attributes (drainage area, elevation, hydrologic unit, etc.).
